Tidal streams in triaxial systems

### 2015-05-12 ms.tex
todos and new nonlinear dynamics section

%TODO:
%The classical way to detect chaotic motion is
%the computation of the Lyapunov exponents. We
%claim that for weakly chaotic solutions, the frequency
%analysis gives information more rapidly
%and more precisely. The reason is that the frequency
%analysis is an instantaneous observation
%of the fundamental frequencies which are essential
%parameters of the dynamical system, while
%the usual computation of Lyapunov exponents
%acts like a mean and needs a very long time span
%to provide reliable results (for references and
%discussion on Lyapunov exponents, see refs. [3,
%12].

% Highlight that stochastic tubes have not been studied much...

@@

% Figure 1
%\begin{figure*}[!h]
%\begin{center}
%\includegraphics[width=0.75\textwidth]{figures/potential.pdf}
%\caption{Rotation curve along $x$, $y$, and $z$ axes for the triaxial, NFW potential } \label{fig:potential}
%\end{center}
%\end{figure*}

@@

For an integrable Hamiltonian, $H$, with $N$ degrees of freedom (dof), the equations of motion are most simply expressed in angle-action coordinates \cite[e.g.,][]{goldstein80}. In these coordinates, the actions, $J_n$, are (isolating) integrals of the motion, and the conjugate angle variables, $\theta_n$, increase linearly with time
\begin{align}
	\dot{\act}_n &= -\frac{\partial H}{\partial \theta_n} = 0\\
	\act_n &= {\rm const.}\\
	\dot{\theta}_n &= \frac{\partial H}{\partial \act_n} = \Omega_n(\act_1,\act_2,...,\act_n)\label{eq:aafreq}\\
	\theta_n(t) &= \Omega_n t + \theta_n(0)
\end{align}
where $i=1,2,...N$, and $\Omega_n$ are the fundamental frequencies of the orbit.  The topology of angle-action space is equivalent to that of the surface of an $N$-dimensional torus, and thus orbits are often referred to in terms of \emph{orbital tori}: each set of actions, $(\act_1,\act_2,...\act_N)$, uniquely ``label'' the torus associated with a given orbit and the angles specify the location on the surface of the $N$-torus. If a global transformation to angle-action coordinates exists, the potential is said to be \emph{integrable}.

Galactic gravitational potentials will almost certainly not be globally integrable. Dark matter halos of galaxies are expected to be triaxial but even if they are close to axisymmetric or spherical, the superposition of components (e.g., disk, bulge, halo) will break global integrability. If a triaxial (three dof) potential is close to integrable such that the Hamiltonian can be written in terms of a small perturbation, $\epsilon$, away from some integrable Hamiltonian, $H_0$,
\begin{equation}
	H(\bs{\theta}, \bs{\act}) = H_0(\bs{\act}) + \epsilon H_1(\bs{\theta}, \bs{\act})
\end{equation}
then a large number of regular orbits will survive the perturbation (by way of the KAM theorem)\footnote{3-vectors are represented with bold symbols, e.g., $\bs{J}=(J_1,J_2,J_3)$.}. The tori that survive will in general be separated by regions of irregular or chaotic motion, and thus any transformations to angle-action coordinates can only be defined locally. Analytic transformations to angle-action coordinates are only known for a handful of simple, integrable potentials. Much effort has gone in to developing general methods for transforming either from position-velocity to angle-action coordinates, $(\bs{x},\bs{v})\rightarrow(\bs{\theta},\bs{J})$ \citep{many}, or vice-versa, $(\bs{\theta},\bs{J})\rightarrow(\bs{x},\bs{v})$ \citep{many}.

The types of motion generic to triaxial systems can be split into several sub-categories (summarized in Figure~\ref{fig:orbit-tree}). Regular motion can be classified by the number of resonance relations obeyed \citep[e.g.,][]{lichtenberg83}: \emph{conditionally periodic} orbits obey no resonance relations, \emph{uni-resonant} orbits obey a single relation of the form $\bs{m}\cdot\bs{\Omega}=0$, and \emph{bi-resonant} orbits obey two resonance relations, $\bs{m}\cdot\bs{\Omega}=0$ and $\bs{n}\cdot\bs{\Omega}=0$, where $\bs{m}$ and $\bs{n}$ are integer vectors.

% Figure 2
%\begin{figure*}[!h]
%\begin{center}
%\includegraphics[width=0.6\textwidth]{figures/orbit-tree.pdf}
%\caption{A rough classification of the types of orbits found in triaxial potentials. Regular orbits are classified by the number of resonance relations obeyed: conditionally periodic (\emph{none}), uni-resonant (\emph{one}), bi-resonant (\emph{two}). Irregular or chaotic orbits are simply separated into two classes: strongly and weakly chaotic, defined by the timescales of interest in a problem. When the characteristic chaotic timescale is much longer than, for example, the age of the universe, the orbit is only weakly chaotic. } \label{fig:orbit-tree}
%\end{center}
%\end{figure*}

Truly resonant orbits are rare \citep{merritt99}, but dynamically important in systems with several degrees of freedom as they determine the structure of orbit-space. Stable, resonant orbits generate families of orbits that may behave or appear similarly, whereas unstable resonances are typically associated with regions of chaos. Not all orbits can be trivially associated with resonances due to the infinitely complex, hierarchical structure of resonances (though the importance or ``size'' of a resonance falls off inversely with the norm of its integer vector, $|\bs{m}|^{-\alpha}$, with $\alpha > 0$). In autonomous, two dof systems, stochastic regions are generally surrounded and separated by regular motion, thus chaotic orbits are confined. In three or more dof the chaotic regions will connect and overlap. This is the origin of Arnold diffusion \citep{arnold64}, which allows chaotic orbits to radically change shape by traversing the web of unstable resonances and are only strictly bounded to their own energy hypersurface. This phenomenon is perhaps easier to visualize in action-space: in two dof systems, if a given set of tori are destroyed due to potential perturbations, an orbit in this now chaotic region is still bounded between the surrounding tori due to energy conservation. In three dimensions, the extra degree of freedom allows chaotic orbits to ``escape'' this confinement.

For small potential perturbations, many regular orbits survive and only small chaotic islands are introduced. As the strength of the perturbation increases, eventually all tori associated with conditionally-periodic motion will be destroyed. Next, the uni-resonant tori are destroyed, and finally the bi-resonant tori --- these are least susceptible to destruction from perturbations \citep[e.g., see][]{valluri98}. However, even orbits that are strictly chaotic may behave regularly over timescales much longer than, for example, the dynamical time of the system. Chaotic orbits can therefore be loosely classified by being either \emph{strongly} or \emph{weakly} chaotic: strongly chaotic orbits behave irregularly over a few ($\sim$10) dynamical times, whereas weakly chaotic orbits behave regularly over this timescale. This rough classification is tied to local measures, such as the Lyapunov time (Section~\ref{sec:lyap}) or frequency diffusion time (Section~\ref{sec:naff}).

These local measures quantify the divergence of infinitesimally nearby orbits or the rate of frequency diffusion along an orbit. For an ensemble of orbits with finite spreads in energy or frequencies (e.g., tidal debris), each orbit will have a slightly different characteristic timescale, but we still expect mixing time of the ensemble to correlate with local measures of chaos of the ``parent'' orbit of the ensemble (Section~\ref{sec:ensemble}). The evolution of tidal debris will depend on the ensemble mixing rate around the parent orbit: chaotic mixing will cause the density of the debris to fall off more quickly than the typical power-law behavior expected for regular orbits \citep[e.g.,][]{merritt96, helmi99}.

@@

%The simplest and perhaps most well-known method for assessing chaotic motion is to analyze the Lyapunov spectrum or maximum Lyapunov exponent (MLE) of an orbit. Defining the vector $\bs{w} = (q_1,...,q_N,p_1,...,p_N)$ for some set of canonical coordinates $(q_i,p_i)$, we can write Hamilton's equations as 
%\begin{equation}
%	\dot{\bs{w}} = \bs{\mathcal{J}} \frac{\partial H}{\partial \bs{w}} = f(\bs{w}) \label{eq:ham}
%\end{equation}
%where $\bs{\mathcal{J}}$ is the $2N \times 2N$ canonical Poisson tensor (also called the symplectic matrix) defined by
%\begin{equation}
%	\bs{\mathcal{J}} = \left( \begin{array}{c:c} 0 & \ident \\ \hdashline -\ident & 0 \end{array} \right)
%\end{equation}
%with $N$-dimensional identity matrices $\ident$. If we consider a nearby phase-space position, $\bs{w}'$, separated from $\bs{w}$ by an infinitesimal deviation, $\delta \bs{w}$, such that $\bs{w}' = \bs{w} + \delta \bs{w}$. We can expand to linear order in the deviation about the parent orbit and write the equations of motion for the deviation as follows
%\begin{align}
%	\dot{\bs{w}}' &= f(\bs{w}')\\
%	\dot{\bs{w}} + \dot{\delta \bs{w}} &= f(\bs{w} +  \delta \bs{w})\\
%	\dot{\bs{w}} + \dot{\delta \bs{w}} &\approx f(\bs{w}) + \jac_{\bs{w}}\cdot\delta \bs{w} + \mathcal{O}(\delta \bs{w}^2)\\
%	\dot{\delta \bs{w}} &\approx \jac_{\bs{w}} \cdot  \delta \bs{w}\label{eq:deviate}
%\end{align}
%where $\jac$ is the Jacobian of Eq.~\ref{eq:ham}, evaluated at the parent orbit (the Hessian of the Hamiltonian). For chaotic orbits, the maximum eigenvalue of the solution matrix to Eq.~\ref{eq:deviate} is positive real, leading to exponential divergence of nearby orbits.

%The simplest and perhaps most well-known method for assessing chaotic motion is to analyze the Lyapunov spectrum or maximum Lyapunov exponent (MLE) of an orbit. Defining the vector $\bs{w} = (q_1,...,q_N,p_1,...,p_N)$ for some set of canonical coordinates $(q_i,p_i)$, and a nearby phase-space position, $\bs{w}'$, separated from $\bs{w}$ by an infinitesimal deviation, $\delta \bs{w}$, such that $\bs{w}' = \bs{w} + \delta \bs{w}$, we can expand to linear order in the deviation about the parent orbit and write the equations of motion for the deviation as 
%\begin{align}
%	\dot{\delta \bs{w}} &\approx \jac_{\bs{w}} \cdot  \delta \bs{w}\label{eq:deviate}
%\end{align}
%where $\jac$ is the Jacobian of Eq.~\ref{eq:ham}, evaluated at the parent orbit (the Hessian of the Hamiltonian). For chaotic orbits, the maximum eigenvalue of the solution matrix to Eq.~\ref{eq:deviate} is positive real, leading to exponential divergence of nearby orbits.

The simplest and perhaps most well-known method for assessing chaotic motion is to analyze the Lyapunov spectrum or maximum Lyapunov exponent (MLE) of an orbit. For some set of canonical coordinates $(q,p)$, we define the vector $w_i = (w_1,...w_{2N}) = (q_1,...,q_N,p_1,...,p_N)$ so we can write Hamilton's equations as 
\begin{equation}
	\dot{w}_i = \mathcal{J}_{ij}\,\frac{\partial H}{\partial w_j} \label{eq:ham}
\end{equation}
where $\mathcal{J}_{ij}$ is the $2N \times 2N$ canonical Poisson tensor (also called the symplectic matrix) defined by
\begin{equation}
	\mathcal{J}_{ij} = \left( \begin{array}{c:c} 0 & \ident \\ \hdashline -\ident & 0 \end{array} \right)
\end{equation}
with $N$-dimensional identity matrices $\ident$. If we consider a nearby phase-space position, $w_i'$, separated from $w_i$ by an infinitesimal deviation, $\delta w_i$, such that $w_i' = w_i + \delta w_i$. We can expand to linear order in the deviation about the parent orbit and write the equations of motion for the deviation as 
\begin{align}
	\dot{\delta w_i} = \mathcal{J}_{ij} \frac{\partial^2 H}{\partial w_j \partial w_k} \delta w_k =  \mathcal{J}_{ij}\, D_{jk} \, \delta w_k \label{eq:deviate}
\end{align}
where $D_{jk}$ is the Hessian matrix evaluated at the parent orbit. For chaotic orbits, the maximum eigenvalue of the solution matrix to Eq.~\ref{eq:deviate} is positive real, leading to exponential divergence of nearby orbits.

@@

[Motivate why we use frequencies vs. actions vs. Lyapunov exponent -- can evaluate small-time movement vs. average?]

@@

% Figure 3
%\begin{figure*}[!h]
%\begin{center}
%\includegraphics[width=0.5\textwidth]{figures/log-freqmap.png}
%\caption{Ratio of cartesian frequencies for a grid of iso-energy box orbits integrated in the potential of Equation~\ref{eq:logpotential}. Stable resonances appear as linear over-densities and unstable resonances appear as linear under-densities. The regularity of the points in this map reflects the input grid of initial conditions. Points that appear to be erratically scattered are chaotic orbits where the frequencies are changing with time.} \label{fig:logfreqs}
%\end{center}
%\end{figure*}

@@

@@

% Figure 4
%\begin{figure*}[!h]
%\begin{center}
%\includegraphics[width=0.9\textwidth]{figures/two-orbits.pdf}
%\caption{ Two orbits integrated in the triaxial potential described in Section~\ref{sec:potential}. Left panels (black) show two consecutive sections of orbit, where the integration time for each section (rows) is set to 100 times the maximum orbital period. Right panels (red) shows the same for the chaotic orbit. The appearance of the orbit in configuration space changes appreciably over timescales of tens of orbital periods. } \label{fig:two-orbits}
%\end{center}
%\end{figure*}

@@

% Figure 5, 6
%\begin{figure*}[!h]
%\begin{center}
%\includegraphics[width=0.62\textwidth]{figures/lyap-orbits.pdf}
%\caption{ FTMLE (Equation~\ref{eq:ftmle}) computed over 1000 orbital periods (the maximum orbital period) for Orbit A (left) and Orbit B (right). For Orbit A, the FTMLE asymptotes towards zero with a characteristic power-law slope. For Orbit B, the FTMLE begins to saturate to a constant; the Lyapunov exponent estimated for this orbit is $\lambda_{\rm max} \approx 0.07 \, {\rm Gyr}^{-1}$ corresponding to a Lyapunov time of $t_\lambda \approx 14 \, {Gyr}$. Over-plotted on both panels are lines (dashed, blue) that show $t^{-0.9}$ behavior.} \label{fig:lyap-orbits}
%\end{center}
%\end{figure*}
%
%\begin{figure*}[!h]
%\begin{center}
%\includegraphics[width=0.62\textwidth]{figures/freq-diff-orbits.pdf}
%\caption{ Difference in frequencies computed for successive sections of 100 orbital periods from the frequencies computed in the initial section for Orbit A (left) and Orbit B (right). The three lines in each panel are for the three fundamental frequencies. Dashed (red) horizontal line shows the estimate of the uncertainty for the frequencies computed for Orbit A. After many hundreds of orbital periods, the frequency deviation is not appreciable indicating again that the orbit is regular or only very weakly chaotic. The frequency deviations computed for Orbit B, however, show a clear increasing trend, indicating that the frequencies are diffusing with time. } \label{fig:fdiff-orbits} % The frequency diffusion rate defined in Equation~\ref{eq:fdrate} measures the initial slope of this trend.
%\end{center}
%\end{figure*}

@@

\section{Test particle ensembles} \label{sec:ensemble}

@@

@@

@@

\subsection{Observable consequences of mixing}

Only effectively looking at debris disrupted from a single pericenter --- debris near progenitor constantly repopulated as more and more stuff disrupts out

Streams should be shorter? How to test this? Length of stream in angle at isodensity contour?

\begin{align}
	E[\rho(\bs{x},t_j)] &= \int \rho(\bs{x},t_j) \, \rho(\bs{x},t_j) \, \dd \bs{x}\\
	&\approx \frac{1}{N}\sum \rho(\bs{x}_i,t_j)
\end{align}
So we can actually estimate the mean density by evaluating at the particles and taking the mean.

@@

\acknowledgements
The authors wish to thank  Andreas K\"upper, Robyn Sanderson, David Hogg, Dan D'Orazio, David Merritt, and the \emph{Stream Team} for useful comments and discussion.
APW is supported by a National Science Foundation Graduate Research Fellowship under Grant No.\ 11-44155. 
This research made use of Astropy, a community-developed core \texttt{Python} package for Astronomy \citep{astropy13}.
This work additionally relied on Columbia University's \emph{Hotfoot} and \emph{Yeti} compute clusters, and we acknowledge the Columbia HPC support staff for assistance.

@@

% Figure 1
\begin{figure*}[!p]
\begin{center}
\includegraphics[width=\textwidth]{figures/potential.pdf}
\caption{Rotation curve along $x$, $y$, and $z$ axes for the triaxial, NFW potential } \label{fig:potential}
\end{center}
\end{figure*}

>>>>

@@

@@

% TODO: move all except chaotic diffusion stuff to appendix?

An orbit in an $N$ degree of freedom (dof) Hamiltonian, $H$, is a set of $2N$ quasi-periodic time series, 
\begin{equation}
(w_1(t),...,w_{2N}(t)) = (q_1(t),...,q_{N}(t),p_1(t),...,p_{N}(t)) \label{eq:coords}
\end{equation}
where $q_i$ and $p_i$ are conjugate coordinates in the sense that
\begin{align}
	\dot{p}_i &= -\frac{\partial H}{\partial q_i}\\
	\dot{q}_i &= \frac{\partial H}{\partial p_i}
\end{align}
for all $t$. The motion in any component, $w_i(t)$, can be represented as a Fourier sum,
\begin{equation}
	w_i(t) = \sum_k^\infty a_k \, e^{i\,\omega_k\,t}; \, (a_k \in \mathbb{C}).
\end{equation}

A \emph{regular orbit} is a set of such time series that can be transformed to a special set of canonical coordinates known as angle-action coordinates. In these coordinates, the position variables are angles, $\boldsymbol{\theta}$, that increase linearly with time with rates set by $N$ constant, fundamental frequencies, $\boldsymbol{\Omega} = (\Omega_1, ..., \Omega_N)$. The frequencies in the Fourier sum of any individual component of motion (the $\omega_k$ in Equation~1) are just linear, integer combinations of the fundamental frequencies, $\boldsymbol{\Omega}$ --- that is, for a regular orbit, any Fourier component frequency may be written
\begin{align}
	\omega_k &= \boldsymbol{n_k} \cdot \boldsymbol{\Omega}; \, (\boldsymbol{n}_k \in \mathbb{Z}^3).
\end{align}
The conjugate momentum coordinates --- the actions,  $\boldsymbol{J}$ --- are constants of motion. Even stronger, the actions are isolating integrals and any pair are in involution such that
\begin{equation}
	[J_i, J_j] = 0
\end{equation}
where $[\cdot,\cdot]$ is the Poisson bracket. This implies that for an $N$ dof system, a regular orbit has $N$ independent constants of motion and the motion is therefore restricted to an $N$-dimensional manifold embedded in the 2$N$ dimensional phase space. The conjugate angle variables define an orthogonal basis, and the topology of angle-action space is therefore the product of $N$ circles, an $N$-torus. Any regular orbit in an $N$ dof Hamiltonian can be understood as motion on the surface of an $N$-torus. Each set of actions, $(J_1,...,J_N)$, uniquely labels a torus, and regular orbits may be referred to in terms of their orbital tori. If a transformation to angle-action variables exists, the Hamiltonian may be written solely in terms of the actions,
\begin{equation}
	H = H(\boldsymbol{J})
\end{equation}

A Hamiltonian or potential is said to be \emph{globally integrable} when the number of isolating integrals of motion is equal to the number of degrees of freedom and a transformation to angle-action coordinates may be done globally --- that is, the transformation to angle-action coordinates may be written as a function of arbitrary phase-space coordinates and the functional form is independent of phase-space position. The condition for global integrability is very restrictive and it seems that the likelihood that a Hamiltonian is globally integrable decreases as the number of degrees of freedom increase \citep[e.g.,][]{lichtenberg83}.

Galactic potentials are almost certainly not globally integrable, but it is useful to understand the orbit structure in integrable systems before extending to more general potentials. Only one globally integrable triaxial mass model is known that generates motion that is qualitatively similar to that seen in Galaxies: the St\"ackel or ``Perfect Ellipsoid'' model \citep[e.g.,][]{kuzmin56, kuzmin73, deZeeuw85, cincotta06}. The four general classes of orbits in this potential are box, inner long-axis tube, outer long-axis tube, and short-axis tube orbits. Unlike the box orbits, the tube orbits preserve a sense of rotation about either the long or short axis.

Over long times, a generic orbit in this potential fills the surface of its 3-torus --- that is, the three fundamental frequencies of the orbit are incommensurable. If there exists a relation of the form
\begin{equation}
	\boldsymbol{n} \cdot \boldsymbol{\Omega} = n\, \Omega_1 + m\, \Omega_2 + l\, \Omega_3 = 0
\end{equation}
then the orbit is a resonant orbit and the effective dimensionality of the orbit is decreased to 2D. We refer to orbits that obey a single resonance relation as \emph{uni-resonant} orbits. If an additional resonance relation exists, the orbit is closed and becomes a 1D curve; we refer to such orbits as \emph{bi-resonant} orbits. The resonant structure of a potential --- the relative importance of particular resonance numbers --- is difficult to compute, but determines the global behavior of orbits in the potential. \cite{cincotta06} analytically compute the resonant structure in action-space for a St\"ackel potential. In frequency-space, the resonances are planes, but in action-space are generic 2D surfaces.

The orbit structure of non-integrable potentials can be qualitatively understood by considering a Hamiltonian that is a small perturbation away from being globally integrable --- that is, a Hamiltonian that may be written
\begin{equation}
	H(\boldsymbol{J}, \boldsymbol{\theta}) = H_0(\boldsymbol{J}) + \epsilon \, H_1(\boldsymbol{J}, \boldsymbol{\theta})
\end{equation}
where $\epsilon$ is a small parameter that determines the perturbation strength \citep[a complete description of perturbation theory applied to nonlinear Hamiltonians is given in][]{lichtenberg83}. When $0 < |\epsilon| \ll 1$, resonant surfaces become ``thick'' resonant layers, within which orbits are qualitatively similar to the parent resonant orbit \citep[e.g.,][]{merritt99} and the stable regions are surrounded by chaotic layers where stochastic motion occurs. Stochastic or chaotic orbits are characterized by apparently random behavior despite being generated by entirely deterministic systems. The stochastic layers form gaps in action space where tori are destroyed. The orbits that exist in these regions therefore can not be represented in angle-action coordinates, as their actions and frequencies evolve with time.

Stochastic motion or the diffusion of an orbit in actions and frequencies may occur either \emph{across} the resonance layers or \emph{along} the stochastic layers (Arnold diffusion). For weakly chaotic orbits, motion across a resonance layer typically occurs with a frequency close to the libration frequency of the nearby stable orbits in the resonance layer. Thus, if the resonance libration frequency is comparable to the orbital frequency, motion across a resonance can modulate the frequency spectrum of an orbit over an orbital time. However, the stochastic layers are often bounded in the direction orthogonal to the resonance by other stable, resonant regions so that the frequencies or actions can't change by large factors (unless there is resonance overlap and the motion is strongly chaotic). The rate of diffusion along stochastic layers via Arnold diffusion depends on the local resonant structure and is hard to predict, however, this has been done analytically for simple potentials \citep[e.g.,][]{chirikov??}. For systems with $N>2$ dof, the stochastic layers connect and form an intricate network of stochasticity known as the Arnold web; the implication of this connectivity is that an orbit may dramatically change its appearance as it navigates through the Arnold web, though the timescales for this to occur are typically many 1000s of orbital periods.

When $\epsilon$ is large, there is no general prediction for the resulting behavior. It seems that more complicated and physically motivated triaxial potential models for galaxies follow the intuition gained from the small-perturbation picture described above, at least for certain parameter choices \citep{valluri, merritt, dehnen, etc.}.

%For an integrable Hamiltonian, $H$, with $N$ degrees of freedom (dof), the equations of motion are most simply expressed in angle-action coordinates \cite[e.g.,][]{goldstein80}. In these coordinates, the actions, $J_n$, are (isolating) integrals of the motion, and the conjugate angle variables, $\theta_n$, increase linearly with time
%\begin{align}
%	\dot{\act}_n &= -\frac{\partial H}{\partial \theta_n} = 0\\
%	\act_n &= {\rm const.}\\
%	\dot{\theta}_n &= \frac{\partial H}{\partial \act_n} = \Omega_n(\act_1,\act_2,...,\act_n)\label{eq:aafreq}\\
%	\theta_n(t) &= \Omega_n t + \theta_n(0)
%\end{align}
%where $i=1,2,...N$, and $\Omega_n$ are the fundamental frequencies of the orbit.  The topology of angle-action space is equivalent to that of the surface of an $N$-dimensional torus, and thus orbits are often referred to in terms of \emph{orbital tori}: each set of actions, $(\act_1,\act_2,...\act_N)$, uniquely ``label'' the torus associated with a given orbit and the angles specify the location on the surface of the $N$-torus. If a global transformation to angle-action coordinates exists, the potential is said to be \emph{integrable}.
%
%Galactic gravitational potentials will almost certainly not be globally integrable. Dark matter halos of galaxies are expected to be triaxial but even if they are close to axisymmetric or spherical, the superposition of components (e.g., disk, bulge, halo) will break global integrability. If a triaxial (three dof) potential is close to integrable such that the Hamiltonian can be written in terms of a small perturbation, $\epsilon$, away from some integrable Hamiltonian, $H_0$,
%\begin{equation}
%	H(\bs{\theta}, \bs{\act}) = H_0(\bs{\act}) + \epsilon H_1(\bs{\theta}, \bs{\act})
%\end{equation}
%then a large number of regular orbits will survive the perturbation (by way of the KAM theorem)\footnote{3-vectors are represented with bold symbols, e.g., $\bs{J}=(J_1,J_2,J_3)$.}. The tori that survive will in general be separated by regions of irregular or chaotic motion, and thus any transformations to angle-action coordinates can only be defined locally. Analytic transformations to angle-action coordinates are only known for a handful of simple, integrable potentials. Much effort has gone in to developing general methods for transforming either from position-velocity to angle-action coordinates, $(\bs{x},\bs{v})\rightarrow(\bs{\theta},\bs{J})$ \citep{many}, or vice-versa, $(\bs{\theta},\bs{J})\rightarrow(\bs{x},\bs{v})$ \citep{many}.
%
%The types of motion generic to triaxial systems can be split into several sub-categories (summarized in Figure~\ref{fig:orbit-tree}). Regular motion can be classified by the number of resonance relations obeyed \citep[e.g.,][]{lichtenberg83}: \emph{conditionally periodic} orbits obey no resonance relations, \emph{uni-resonant} orbits obey a single relation of the form $\bs{m}\cdot\bs{\Omega}=0$, and \emph{bi-resonant} orbits obey two resonance relations, $\bs{m}\cdot\bs{\Omega}=0$ and $\bs{n}\cdot\bs{\Omega}=0$, where $\bs{m}$ and $\bs{n}$ are integer vectors.
%
%% Figure 2
%%\begin{figure*}[!h]
%%\begin{center}
%%\includegraphics[width=0.6\textwidth]{figures/orbit-tree.pdf}
%%\caption{A rough classification of the types of orbits found in triaxial potentials. Regular orbits are classified by the number of resonance relations obeyed: conditionally periodic (\emph{none}), uni-resonant (\emph{one}), bi-resonant (\emph{two}). Irregular or chaotic orbits are simply separated into two classes: strongly and weakly chaotic, defined by the timescales of interest in a problem. When the characteristic chaotic timescale is much longer than, for example, the age of the universe, the orbit is only weakly chaotic. } \label{fig:orbit-tree}
%%\end{center}
%%\end{figure*}
%
%Truly resonant orbits are rare \citep{merritt99}, but dynamically important in systems with several degrees of freedom as they determine the structure of orbit-space. Stable, resonant orbits generate families of orbits that may behave or appear similarly, whereas unstable resonances are typically associated with regions of chaos. Not all orbits can be trivially associated with resonances due to the infinitely complex, hierarchical structure of resonances (though the importance or ``size'' of a resonance falls off inversely with the norm of its integer vector, $|\bs{m}|^{-\alpha}$, with $\alpha > 0$). In autonomous, two dof systems, stochastic regions are generally surrounded and separated by regular motion, thus chaotic orbits are confined. In three or more dof the chaotic regions will connect and overlap. This is the origin of Arnold diffusion \citep{arnold64}, which allows chaotic orbits to radically change shape by traversing the web of unstable resonances and are only strictly bounded to their own energy hypersurface. This phenomenon is perhaps easier to visualize in action-space: in two dof systems, if a given set of tori are destroyed due to potential perturbations, an orbit in this now chaotic region is still bounded between the surrounding tori due to energy conservation. In three dimensions, the extra degree of freedom allows chaotic orbits to ``escape'' this confinement.
%
%For small potential perturbations, many regular orbits survive and only small chaotic islands are introduced. As the strength of the perturbation increases, eventually all tori associated with conditionally-periodic motion will be destroyed. Next, the uni-resonant tori are destroyed, and finally the bi-resonant tori --- these are least susceptible to destruction from perturbations \citep[e.g., see][]{valluri98}. However, even orbits that are strictly chaotic may behave regularly over timescales much longer than, for example, the dynamical time of the system. Chaotic orbits can therefore be loosely classified by being either \emph{strongly} or \emph{weakly} chaotic: strongly chaotic orbits behave irregularly over a few ($\sim$10) dynamical times, whereas weakly chaotic orbits behave regularly over this timescale. This rough classification is tied to local measures, such as the Lyapunov time (Section~\ref{sec:lyap}) or frequency diffusion time (Section~\ref{sec:naff}). 
%
%These local measures quantify the divergence of infinitesimally nearby orbits or the rate of frequency diffusion along an orbit. For an ensemble of orbits with finite spreads in energy or frequencies (e.g., tidal debris), each orbit will have a slightly different characteristic timescale, but we still expect mixing time of the ensemble to correlate with local measures of chaos of the ``parent'' orbit of the ensemble (Section~\ref{sec:ensemble}). The evolution of tidal debris will depend on the ensemble mixing rate around the parent orbit: chaotic mixing will cause the density of the debris to fall off more quickly than the typical power-law behavior expected for regular orbits \citep[e.g.,][]{merritt96, helmi99}. 
%

@@

% TODO: move most of this to appendix? leave only discussion?

The simplest and perhaps most well-known method for assessing chaotic motion is to analyze the Lyapunov spectrum or maximum Lyapunov exponent (MLE) of an orbit. Using our definition of $w$ from Equation~\ref{eq:coords}, we can write Hamilton's equations as 
\begin{equation}
	\dot{w}_i = \mathcal{J}_{ij}\,\frac{\partial H}{\partial w_j} \label{eq:ham}
\end{equation}
where $\mathcal{J}_{ij}$ is the $2N \times 2N$ canonical Poisson tensor (also called the symplectic matrix) defined by
\begin{equation}
	\mathcal{J}_{ij} = \left( \begin{array}{c:c} 0 & \ident \\ \hdashline -\ident & 0 \end{array} \right)
\end{equation}
with $N$-dimensional identity matrices $\ident$. If we consider a nearby phase-space position, $w_i'$, separated from $w_i$ by an infinitesimal deviation, $\delta w_i$, such that $w_i' = w_i + \delta w_i$. We can expand to linear order in the deviation about the parent orbit and write the equations of motion for the deviation as 
\begin{align}
	\dot{\delta w_i} = \mathcal{J}_{ij} \frac{\partial^2 H}{\partial w_j \partial w_k} \delta w_k =  \mathcal{J}_{ij}\, D_{jk} \, \delta w_k \label{eq:deviate}
\end{align}
where $D_{jk}$ is the Hessian matrix evaluated at the parent orbit. For chaotic orbits, the maximum eigenvalue of the solution matrix to Eq.~\ref{eq:deviate} is positive real, leading to exponential divergence of nearby orbits.

@@

% TODO: move most to appendix, leave only discussion?

@@

@@

[Can see lots of cool structure, but rates are tiny! Why does this manifest itself over 10-20 periods?]

@@

@@

@@

@@

% TODO: with 3 N-body simulations of regular, weakly chaotic, strongly chaotic, show that ensembles reproduce qualitative behavior

@@

\subsection{Frequency variance}\label{sec:freqvar}

Both the Lyapunov exponent and the frequency diffusion rate measure mean, long-term rates of chaotic diffusion. As discussed in Section~\ref{sec:nldreview}, frequency diffusion may occur either across a resonance layer or along a stochastic layer. If a weakly chaotic orbit is confined (by other nearby, non-overlapping resonances or stable regions), the mean drift of an orbit in frequency space will be small if computed over timescales long compared to the orbital time but short compared to the Arnold diffusion time. For the potential considered in this paper, these are precisely the timescales for which the frequency diffusion rate of the previous section were computed ($\sim$100's of orbital periods). This window length was chosen in order to have a strong enough signal in the frequency spectrum of chaotic orbits to measure the most significant frequencies using NAFF.

In this section, we instead wish to resolve the short-time behavior of the frequency diffusion (corresponding to motion across resonance layers). For each orbit, we now instead use a rolling window to analyze the evolution of the frequency spectrum on shorter times. We use a window with a width of 50 orbital periods and shift the window by half an orbital period between each calculation of the most significant frequencies. [Why this window width? Stride?]

@@

@@

\acknowledgements
The authors wish to thank ... Robyn Sanderson, David Hogg, Dan D'Orazio, David Merritt, and the \emph{Stream Team} for useful comments and discussion.
APW is supported by a National Science Foundation Graduate Research Fellowship under Grant No.\ 11-44155. 
This research made use of Astropy, a community-developed core \texttt{Python} package for Astronomy \citep{astropy13}.
This work additionally relied on Columbia University's \emph{Hotfoot} and \emph{Yeti} compute clusters, and we acknowledge the Columbia HPC support staff for assistance.

@@

% Figure 1
%\begin{figure*}[!p]
%\begin{center}
%\includegraphics[width=\textwidth]{figures/potential.pdf}
%\caption{Rotation curve along $x$, $y$, and $z$ axes for the triaxial, NFW potential } \label{fig:potential}
%\end{center}
%\end{figure*}
